Motivating Exploration in Reinforcement Learning

### 2020-04-17 paper/main.tex
Final sweep of report

Exploration is something that comes naturally to humans, almost as if there exists an innate sense of curiosity, craving to experience something new, something yet to be seen. Jean Piaget, a Swiss developmental psychologist, asserted that it is in human nature to be curious, inquisitive, and interested in novelty as they develop and acquire knowledge about their environment \cite{piaget}.Through observing human behaviour, it is clear that the ability to explore a domain is critical to the process of discovery and learning, regardless of the task being considered.

@@

\subsection{Advantage Actor-Critic (A2C)}
Our baseline is built around A2C, which is an on-policy learning algorithm. The core decision making of the agent stems from the policy network ($\pi$), which is also referred to as the \textit{Actor}. The policy network is learned and its weights ($\theta_\pi$) are updated via the standard policy gradient equation, wherein the rewards $r_t$ for a trajectory $t$ are weighted by the negative log likelihood of that trajectory. Minimizing this loss $\mathcal{L}_P$ is equivalent to updating $\pi$ such that the probability of high reward trajectories are maximized.
\begin{equation}
\mathcal{L}_{P} = \sum_{t} -\log\pi(s_t; \theta_\pi) \times r_t\label{eq:policy}
\end{equation}
A2C also uses the value network ($V$) or \textit{critic} which predicts the accumulated, discounted rewards $R_i$ over the episode timesteps $i$. The advantages can be computed as $A_i = R_i - V_i$, where $V_i$ are the predictions from $V$, and are used in place of the actual rewards in the policy update Eq. \ref{eq:policy}, reducing the variance of $\mathcal{L}_P$, which is a common downfall of on-policy methods. We train the value network $V$ with update Eq. \ref{eq:value}.
\begin{equation}
\mathcal{L}_V = \frac{1}{n}\sum_i (R_i - V_i)^2\label{eq:value}
\end{equation}
The base A2C algorithm does not explicitly seek random exploration as it learns to decide on the action at a https://arxiv.org/pdf/1601.06733.pdf state purely from the policy update. The most commonly used method to introduce randomness to the agent is through maximizing the entropy of the next-action sampling distribution. This distribution is categorical with probabilities defined by the policy network logits.
\begin{equation}\label{eq:entropy}
\mathcal{L}_E = \sum_a (-\log \pi_a \times \mu_\pi)
\end{equation}
Instead of minizing the loss function in Eq. \ref{eq:entropy}, we maximize the value so that the individual probabilities of the sampling distribution $\pi_a$ are as similar as possible; thus, inducing a degree of randomness in the policy network's state to action mapping.

\subsection{Intrinsic Curiosity}
Entropy encourages the agent to explore but in a purely random manner, which has been shown to produce exceptional results (refer to experiments). However, we also compare with a more targeted approach by leveraging the Intrinsic Curiosity Module (ICM) introduced in \cite{curiosity}, which provides an intrinsic reward as a learning signal. ICM consists of two linear networks referred to as the forward and inverse models. The forward model $f_{Fwd}$ is trained to predict the next state ($\phi(s_{t+1})$) given the current state ($\phi(s_t)$) and the action ($a_t$) decided by the policy net. Note that we use $\phi$ to denote a feature extractor network used to encode the raw states.
\begin{equation}\label{eq:intrinsic}
r_i = \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
\begin{equation}\label{eq:reward}
r = r_e + \beta r_i
\end{equation}
The new intrinsic reward $r_i$ is defined as the mean-squared-error of the predicted next state and the actual next state, which is then combined with the extrinsic reward to obtain the total reward used in the policy update Eq. \ref{eq:policy}. Thus, our agent learns to maximize the likelihood of following trajectories that maximize the intrinsic reward, which means that the agent is encouraged to enter new, unseen states that $f_{Fwd}$ is unable to predict accurately. The forward model itself is trained using the loss below.
\begin{equation}\label{eq:forward}
\mathcal{L}_{Fwd} = \frac{1}{n} \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
Further, the purpose of the inverse model $f_{Inv}$ is to train the feature extractor $\phi$ to encode observed features from the raw state that are most impacted by the agent's actions. Specifically, the inverse model takes the current and next states and infers the corresponding action taken: $f_{Inv}(\phi(s_t), \phi(s_{t+1})) = \hat{p}$, where $\hat{p}$ is a vector of probabilities for each possible action. This is trained with the following,
\begin{equation}\label{eq:inverse}
\mathcal{L}_{Inv} = -\log \Big(\frac{e^{\hat{p}[a_t]}}{\sum_j e^{\hat{p}[j]}}\Big) = -\hat{p}[a_t] + \log\Big(\sum_j e^{\hat{p}[j]}\Big)
\end{equation}
The described curiosity model is incorporated into the base A2C model and can be trained jointly:
\begin{equation}
\mathcal{L} = \mathcal{L}_P + \lambda_{V}\mathcal{L}_V - \lambda_{E}\mathcal{L}_E + \mathcal{L}_{Fwd} + \mathcal{L}_{Inv}
\end{equation}
where we use $\lambda_V=0.5$ and $\lambda_E=0.02$.

@@

\subsubsection{Dense Extrinsic Reward}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/breakout_rewards}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/seaquest_rewards}
\end{subfigure}
\caption{Here we compare the three A2C variants in the \texttt{Breakout} (left) and \texttt{Seaquest} (right) environments. Rewards are again cumulative over an entire episode.}
\label{fig:seaquest-breakout-reward}
\end{figure}

We refer to \texttt{Breakout} and \texttt{Seaquest} as our dense reward environments. Specifically, \texttt{Seaquest} is most dense in that positive rewards are given for each ``kill'', which occurs very frequently due to the sheer number of enemies and the fact that the player's submarine is always firing projectiles at high velocity. In fact, positive rewards are already obtained in the early steps of training (as can be seen in Fig. \ref{fig:seaquest-breakout-reward}). \texttt{Breakout} provides a slightly sparser reward but is still denser than that of \texttt{Pong}; rewards are given for each brick broken which occurs every time the agent is able to return the ball. This means that in early training, when the agent is unable to consistently return the ball, there is no reward signal and so training performance is heavily dependent on an exploration factor.

@@

\bibliography{main}
\bibliographystyle{ieeetr}

>>>>

Exploration is something that comes naturally to humans, almost as if there exists an innate sense of curiosity, craving to experience something new, something yet to be seen. Jean Piaget, a Swiss developmental psychologist, asserted that it is in human nature to be curious, inquisitive, and interested in novelty as they develop and acquire knowledge about their environment \cite{piaget}. Through observing human behaviour, it is clear that the ability to explore a domain is critical to the process of discovery and learning, regardless of the task being considered.

@@

\subsection{Advantage Actor-Critic (A2C)}
Our baseline is built around A2C, which is an on-policy learning algorithm. The core decision making of the agent stems from the policy network ($\pi$), which is also referred to as the \textit{Actor}. The policy network is learned and its weights ($\theta_\pi$) are updated via the standard policy gradient equation, wherein the rewards $r_t$ for a trajectory $t$ are weighted by the negative log likelihood of that trajectory. Minimizing this loss $\mathcal{L}_P$ is equivalent to updating $\pi$ such that the probability of high reward trajectories are maximized.
\begin{equation}
\mathcal{L}_{P} = \sum_{t} -\log\pi(s_t; \theta_\pi) \times r_t\label{eq:policy}
\end{equation}
A2C also uses the value network ($V$) or \textit{critic} which predicts the accumulated, discounted rewards $R_i$ over the episode timesteps $i$. The advantages can be computed as $A_i = R_i - V_i$, where $V_i$ are the predictions from $V$, and are used in place of the actual rewards in the policy update Eq. \ref{eq:policy}, reducing the variance of $\mathcal{L}_P$, which is a common downfall of on-policy methods. We train the value network $V$ with update Eq. \ref{eq:value}.
\begin{equation}
\mathcal{L}_V = \frac{1}{n}\sum_i (R_i - V_i)^2\label{eq:value}
\end{equation}
The base A2C algorithm does not explicitly seek random exploration as it learns to decide on the action at a state purely from the policy update. The most commonly used method to introduce randomness to the agent is through maximizing the entropy $E$, shown in Fig. \ref{eq:entropy}, of the next-action sampling distribution. This distribution is categorical with probabilities defined by the policy network logits.
\begin{equation}\label{eq:entropy}
E = \sum_a (-\log \pi_a \times \mu_\pi)
\end{equation}
Instead of minizing the loss function in Eq. \ref{eq:entropy}, we maximize the value so that the individual probabilities of the sampling distribution $\pi_a$ are as similar as possible; thus, inducing a degree of randomness in the policy network's state to action mapping.

\subsection{Intrinsic Curiosity}
Entropy encourages the agent to explore but in a purely random manner, which has been shown to produce exceptional results (refer to experiments). However, we also compare with a more targeted approach by leveraging the Intrinsic Curiosity Module (ICM) introduced in \cite{curiosity}, which provides an intrinsic reward as a learning signal. ICM consists of two linear networks referred to as the forward and inverse models. The forward model $f_{Fwd}$ is trained to predict the next state ($\phi(s_{t+1})$) given the current state ($\phi(s_t)$) and the action ($a_t$) decided by the policy net. Note that we use $\phi$ to denote a feature extractor network used to encode the raw states.
\begin{equation}\label{eq:intrinsic}
r_i = \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
\begin{equation}\label{eq:reward}
r = r_e + \beta r_i
\end{equation}
The new intrinsic reward $r_i$ is defined as the mean-squared-error of the predicted next state and the actual next state, which is then combined with the extrinsic reward to obtain the total reward used in the policy update Eq. \ref{eq:policy}. Thus, our agent learns to maximize the likelihood of following trajectories that maximize the intrinsic reward, which means that the agent is encouraged to enter new, unseen states that $f_{Fwd}$ is unable to predict accurately. The forward model itself is trained using the loss below.
\begin{equation}\label{eq:forward}
\mathcal{L}_{Fwd} = \frac{1}{n} \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
Further, the purpose of the inverse model $f_{Inv}$ is to train the feature extractor $\phi$ to encode observed features from the raw state that are most impacted by the agent's actions. Specifically, the inverse model takes the current and next states and infers the corresponding action taken: $f_{Inv}(\phi(s_t), \phi(s_{t+1})) = \hat{p}$, where $\hat{p}$ is a vector of probabilities for each possible action. This is trained with the following,
\begin{equation}\label{eq:inverse}
\mathcal{L}_{Inv} = -\log \Big(\frac{e^{\hat{p}[a_t]}}{\sum_j e^{\hat{p}[j]}}\Big) = -\hat{p}[a_t] + \log\Big(\sum_j e^{\hat{p}[j]}\Big)
\end{equation}
The described curiosity model is incorporated into the base A2C model and can be trained jointly:
\begin{equation}
\mathcal{L} = \mathcal{L}_P + \lambda_{V}\mathcal{L}_V - \lambda_{E}E + \mathcal{L}_{Fwd} + \mathcal{L}_{Inv}
\end{equation}
where we use $\lambda_V=0.5$ and $\lambda_E=0.02$.

@@

\subsubsection{Dense Extrinsic Reward}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/breakout_rewards}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/seaquest_rewards}
\end{subfigure}
\caption{Here we compare the three A2C variants in the \texttt{Breakout} (left) and \texttt{Seaquest} (right) environments. Rewards are again cumulative over an entire episode. We note that the performance of the curious learner (cyan) is worse at times than the A2C + Entropy model (yellow), which is perhaps due to the lack of hyper-parameter tuning and the density of the extrinsic reward in these environments.}
\label{fig:seaquest-breakout-reward}
\end{figure}

We refer to \texttt{Breakout} and \texttt{Seaquest} as our dense reward environments. Specifically, \texttt{Seaquest} is most dense in that positive rewards are given for each ``kill'', which occurs very frequently due to the sheer number of enemies and the fact that the player's submarine is always firing projectiles at high velocity. In fact, positive rewards are already obtained in the early steps of training (as can be seen in the right of Fig. \ref{fig:seaquest-breakout-reward}). \texttt{Breakout} provides a slightly sparser reward but is still denser than that of \texttt{Pong}; rewards are given for each brick broken which occurs every time the agent is able to return the ball. This means that in early training, when the agent is unable to consistently return the ball, there is no reward signal and so training performance is heavily dependent on an exploration factor.

@@

\newpage
\bibliography{main}
\bibliographystyle{ieeetr}
